My title

### 2024-03-17 paper/paper.qmd
Updated data, model section

## Data Sources
The Alberta Government created the dataset.The official website of the Government of Alberta provides a single platform for services and information pertaining to Alberta, Canada. It provides information on government news, jobs, social services, health, and education, making it an essential tool for both locals and tourists to be updated about provincial issues.

## Variables
Order by total number of deaths and a ranking of the top 30 causes of death in Alberta each year.Our table lists the top eight causes of death in Alberta in 2022. Examine each variable in detail:
*Year*: This denotes the data gathering year, which for all entries is 2022.
*Cause*: This represents the medical condition or event that led to death. The causes listed are shown in @fig-dataset-overview:
`Organic dementia`
`All other forms of chronic...`, 
`Other ill-defined and unknown...`, 
`COVID-19, virus identified`, 
`Malignant neoplasms of the trachea, bronchus, and lung`, 
`Acute myocardial infarction`, 
`Accidental poisoning by and...`, 
`Other chronic obstructive pulmonary diseases`, 
*Ranking*: This is a ranking by the number of deaths caused by each disease, with 1 being the highest.
*Deaths*: The number of deaths attributed to each cause.
*Year*: Indicates the number of years in which data was collected for that reason.

@@

## Data Description
It accurately reflects health issues specific to the region, but this information is somewhat limited by location. Health policies vary from country to country, and in some regions and countries, health coverage may be less developed than in others, and low-income groups may not be able to afford basic medical care, which increases their risk of chronic diseases and other health problems. The main types of employment in a city also have an impact on health; In an industrial city, some jobs may require exposure to hazardous substances or strenuous physical labor, which increases the likelihood of developing occupational diseases. Risks associated with the workplace may lead to an increase in diseases such as cardiovascular and respiratory diseases. In addition, communities with higher levels of education and economic prosperity generally have stronger social support systems, which are critical for disease prevention.

The demographics of a society also have a significant impact on cause-of-death data. Age distribution is important because certain age groups are more susceptible to certain diseases. For example, young children may be more susceptible to infectious diseases, while older people may be more susceptible to degenerative and chronic diseases. Given that certain medical conditions are more common in one sex than the other - men may have higher rates of heart disease, while women may have higher rates of certain cancers - the sex ratio may also have an impact on causes of death. In addition, differences in the ethnic makeup of the population may be related to lifestyle choices, genetic makeup, and disease risk. For example, genetic factors may contribute to the higher incidence of certain chronic diseases in particular ethnic groups. Elucidating how resources are allocated to public health problems therefore requires a thorough understanding and study of population characteristics.

We can also see that in the data table @fig-dataset-overview for 2022, COVID-19 ranks fourth, with 1,547 deaths. Alberta's cause-of-death data proves that the coronavirus pandemic is one of the most serious public health crises of the early 21st century, with global implications. This figure not only speaks to the outbreak's death rate, but also suggests that the region's health care system may be under strain. The COVID-19 pandemic has disrupted the distribution of deaths from previously common causes, which may include some chronic diseases that have long held the top spot, such as cancer and heart disease, even though it was not the leading cause of death in the years shown in the table.

As large datasets are difficult to observe, this report will analyze only specific aspects. The original dataset contains eight different causes of death. Only the first five causes will be analyzed in this paper.

For simplicity we restrict ourselves to the five most common causes of death in 2022 of those that have been present every year.
```{r}
#| label: tbl-top-five
#| tbl-cap: Top Five Leading causes of death in 2022
#| echo: false
#| warning: false
# Look up the five most common causes of death in 2022
alberta_cod_top_five <-
  alberta_cod |>
  filter(
     calendar_year == 2022
  ) |>
  slice_max(order_by = desc(ranking), n = 5) |>
  pull(cause)

@@

head(alberta_cod_top_five)
```

@@

@@

```{r}
#| label: tbl-mean-var
#| tbl-cap: Comparison of Mean and Variance of Total Death in from 2001 to 2022
#| echo: false
#| warning: false
# Calculate the mean and variance
summary_stats <- alberta_cod %>%
  summarise(
    Mean = mean(total_deaths, na.rm = TRUE),  # Calculate mean, remove NA values if any
    Variance = var(total_deaths, na.rm = TRUE) # Calculate variance, remove NA values if any
  )

# Create a table
kable(summary_stats, caption = "Comparison of Mean and Variance of Total Deaths")

@@

We created a figure @fig-ppcheckandposteriorvsprior that includes both the Poisson and negative binomial models. For the Poisson model, the actual observed values y are represented in black. The predicted values $y_{Rep}$ are shown in light blue, which are replicated samples generated based on the posterior distribution of model parameters.

@@

For the negative binomial models, the actual observed values y are also represented in black, and the predicted values yrep are likewise shown in light blue, generated based on the posterior distribution of the model parameters. Compared to the Poisson model, the predictive distributions generated by the negative binomial model appear to be more concentrated, meaning they are more closely clustered around the actual observed values at higher counts, indicating that the negative binomial model may be more suitable for these data.
From the figure, it can be observed that the predictive distributions of the negative binomial model match the actual observed values better than those of the Poisson model. In the Poisson model's figure, the predicted values align well with the actual values near the peak in the middle but are wider in the tail distribution, which may mean that the Poisson model struggles with extreme values. In contrast, in the figure for the negative binomial model, the predicted values seem to follow the actual observed values more closely across the entire range, typically indicating it can better handle the overdispersion of the data.

```{r}
#| echo: false
#| eval: true
#| message: true
#| warning: true
#| label: fig-LOOIC
#| fig-cap: "Comparing the LOO-CV scores"

poisson <- loo(cause_of_death_alberta_poisson, cores = 1)
neg_binomial <- loo(cause_of_death_alberta_neg_binomial, cores = 1)

loo_compare(poisson, neg_binomial)
```

@@

\appendix

# Appendix {-}

# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...

>>>>

## Data Sources
Our research team conducted an in-depth analysis of mortality trends in Alberta, choosing a dataset published on the Alberta Open Government Portal as the basis for their study. This dataset was made available to the public beginning in 2015 and was last updated in 2023, and is designed to promote public engagement by providing transparent access to key health statistics.This dataset provides a comprehensive view of deaths across the province, covering the top 30 leading causes of death in each annual count. This dataset was chosen because it provides key statistics on the number of deaths and mortality rates. These statistics cover data from chronic diseases to unintended injuries and even public health emergencies, allowing us to accurately track trends in key causes of death. We chose to rely on this dataset primarily because it is sourced from the official government of Alberta, which provides an authoritative data base for our study. The endorsement of a government department ensures the accuracy of the data, which is a key prerequisite for conducting scientific research. In addition, the high degree of fit of the dataset with our research objectives was an important reason for its selection. Our study aimed to gain insight into the key factors that influence mortality in Alberta and to explore possible public health interventions. By analysing this data, we are able to identify priority areas for health interventions and provide a scientific basis for policy makers.

## Features
Order by total number of deaths and a ranking of the top 30 causes of death in Alberta each year.Our table lists the top eight causes of death in Alberta in 2022. Examine each variable in detail:
*Year*: This denotes the data gathering year, which for all entries is 2022.
*Cause*: This represents the medical condition or event that led to death. The causes listed are shown in @fig-dataset-overview:
`Organic dementia`
`All other forms of chronic...`, 
`Other ill-defined and unknown...`, 
`COVID-19, virus identified`, 
`Malignant neoplasms of the trachea, bronchus, and lung`, 
`Acute myocardial infarction`, 
`Accidental poisoning by and...`, 
`Other chronic obstructive pulmonary diseases`, 
*Ranking*: This is a ranking by the number of deaths caused by each disease, with 1 being the highest.
*Deaths*: The number of deaths attributed to each cause.
*Years*: Indicates the number of years in which data was collected for that reason.

@@

## Data Method
In the preliminary phase, we worked on organizing and harmonizing the naming of variables in the dataset to enhance its readability and efficiency of analysis. In addition, we summarized the data in order to quantify the incidence of various causes of death, thus making the dataset more streamlined for more in-depth analysis. To avoid missing years in the dataset from affecting the results of the analysis, we decided to delete all records containing these missing values.We then prioritized the top 8 causes of death by 2022 based on their number of deaths. Particular attention was paid to the top 5 leading causes of death, which is at a critical time when the world is facing a pandemic virus, and the data allow us to understand the impact of the new crown on mortality.Our mortality data are consistent with the characteristic of occurring independently with equal means and variances. These mortality trends are fitted by Poisson and negative binomial statistical models because of the suitability of both models for working with count data. A method for assessing the effect of predictor variables on mortality changes is provided, with exponential coefficients that visually represent the change in mortality risk that may result from each unit change in a predictor variable.

```{r}
#| label: tbl-top-five
#| echo: false
#| warning: false
# Look up the five most common causes of death in 2022
alberta_cod_top_five <-
  alberta_cod |>
  filter(
     calendar_year == 2022
  ) |>
  slice_max(order_by = desc(ranking), n = 5) |>
  pull(cause)

@@

alberta_cod_top_five <- data.frame(CauseOfDeath = alberta_cod_top_five)

# New names for the causes of death
new_names <- c("Organic dementia",
               "All other forms of chronic ischemic heart disease",
               "Other ill-defined and unknown causes of mortality",
               "COVID-19, virus identified",
               "Malignant neoplasms of trachea, bronchus and lung")

# Assign these new names to the CauseOfDeath column
alberta_cod_top_five$CauseOfDeath <- new_names

# View the updated data frame

# Create the table with complete variable names
kable(alberta_cod_top_five, 
      caption = "Top Five Causes of Death in Alberta for 2022")

We can also see that in the data table for 2022, COVID-19 ranks fourth, with 1,547 deaths. Alberta's cause-of-death data proves that the coronavirus pandemic is one of the most serious public health crises of the early 21st century, with global implications. This figure not only speaks to the outbreak's death rate, but also suggests that the region's health care system may be under strain. The COVID-19 pandemic has disrupted the distribution of deaths from previously common causes, which may include some chronic diseases that have long held the top spot, such as cancer and heart disease, even though it was not the leading cause of death in the years shown in the table.

@@

```{r}
#| label: tbl-mean-var
#| echo: false
#| warning: false
# Calculate the mean and variance
summary_stats <- alberta_cod %>%
  summarise(
    Mean = mean(total_deaths, na.rm = TRUE),  # Calculate mean, remove NA values if any
    Variance = var(total_deaths, na.rm = TRUE) # Calculate variance, remove NA values if any
  )

# Create a table
kable(summary_stats, caption = "Comparison of Mean and Variance of Total Deaths in Alberta from 2001 to 2022")

@@

@@

We created @fig-ppcheckandposteriorvsprior that includes both the Poisson and negative binomial models. For the Poisson model, the actual observed values y are represented in black. The predicted values *$y_{Rep}$* are shown in light blue, which are replicated samples generated based on the posterior distribution of model parameters.

@@

For the negative binomial models, the actual observed values `y` are also represented in black, and the predicted values yrep are likewise shown in light blue, generated based on the posterior distribution of the model parameters. Compared to the Poisson model, the predictive distributions generated by the negative binomial model appear to be more concentrated, meaning they are more closely clustered around the actual observed values at higher counts, indicating that the negative binomial model may be more suitable for these data.
From the figure, it can be observed that the predictive distributions of the negative binomial model match the actual observed values better than those of the Poisson model. In the Poisson model's figure, the predicted values align well with the actual values near the peak in the middle but are wider in the tail distribution, which may mean that the Poisson model struggles with extreme values. In contrast, in the figure for the negative binomial model, the predicted values seem to follow the actual observed values more closely across the entire range, typically indicating it can better handle the overdispersion of the data.

```{r}
#| echo: false
#| eval: true
#| message: true
#| warning: false
#| label: tbl-LOOIC

# Perform LOO-CV (Leave-One-Out Cross-Validation) using the loo package
poisson <- loo(cause_of_death_alberta_poisson, cores = 1)
neg_binomial <- loo(cause_of_death_alberta_neg_binomial, cores = 1)

# Compare the models
comparison <- loo_compare(poisson, neg_binomial)

# Convert the comparison to a data frame for better handling (optional)
comparison_df <- as.data.frame(comparison)

# Select only the columns for ELPD Diff and SE Diff
selected_comparison_df <- comparison_df[, c("elpd_diff", "se_diff")]

# Set the row names to more descriptive labels
rownames(selected_comparison_df) <- c("Negative Binomial Model", "Poisson Model")

# Optionally, rename the columns for clarity
names(selected_comparison_df) <- c("Difference in Expected Log Predictive Density (ELPD Difference)", "Standard Error of ELPD Difference (SE Difference)")

# Now create the table with the updated column names and selected columns
kable(selected_comparison_df, caption = "LOO-CV Model Comparison: ELPD Difference and SE Difference", align = 'c')
```

@@
